---
title: "Spatial Analysis of Cardiovascular Diseases in the Northeastern United States"
author: "Michael Ontiveros"
output: 
    html_document:
        theme: cosmo 
        toc: true
        toc_float: true
        highlight: tango
        number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
set.seed(9999)
```

# Details

+ Use the US Census Bureau's definition of the Northeast: [click here](https://www.wikiwand.com/en/Northeastern_United_States)
+ Fix the hole with missing data in our shapefile. Look at prior poster for relevant code. 
    + Shannon County (GEOID 46113) is now known as Oglala Lakota (GEOID 46102).
    + Wade Hamptom (GEOID 02270) is now known as Kusilvak (GEOID 02158).
+ Use 2017 data ACS CDC WONDER
+ check for correlation between covariates 
+ Next year, the US Census website—and not American FactFinder—will host the demographic data used for this case study.
+ Consider wrangling FIPS excel sheet found online
+ Reorder urbanization categories to correctly reflect "rural-ness."
+ Use head() or glimpse according to dim() output

# Motivation

## Background

Cardiovascular diseases are currently the leading cause of death in the United States. Much is already known about the causes of most cardiovascular diseases. We intend to explore the environmental characteristics related to the mortality of these diseases. 

## Analysis goal

We will identify county-level characteristics that are associated with cardiovascular disease mortality in the Northeastern United States, giving particular attention to the potential effects of neighboring counties. 

## Learning objectives

The learning objectives include data wrangling of spatial data, spatial smoothing, and the creation and interpretation of spatially lagged models.

## Libraries

```{r}
library(MASS)
library(reshape2)
library(tidyverse)
library(readxl)
library(sf)
library(tmap)
library(viridis)
library(spdep) #for neighborhood matrix, correlogram
library(car) #for variance inflation factor calculation
```

# What is the data?

We will be using data from two sources. 

We will be obtaining demographic and geographic information from the US Census Bureau.

Our analysis will use demographic information from the American Community Survey, which functions similarly to the US Census, but is conducted yearly. 

Two map files for this analysis will be considered. You can obtain these map files [here](https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.2017.html) and [here](https://www.census.gov/cgi-bin/geo/shapefiles/index.php). We will use the *2017 State, County, Minor Civil Division, and Incorporated Place FIPS Codes* provided [here](https://www.census.gov/geographies/reference-files/2017/demo/popest/2017-fips.html) to make identification of states and counties more practical in our analysis. 

Our second source of data is underlying cause of death data from CDC WONDER. Simply put, each death record contains separate sections for the underlying cause of death and all multiple causes of death. The CDC compiles information from these death records to create the respective datasets.

The way we define deaths in our analysis is important to consider. We will be defining deaths with a cardiovascular disease listed as the underlying cause of death as a death as a result of a cardiovascular disease.

Not all deaths have a single, direct cause. For that very reason, all multiple causes of death are listed on death records. We've defined our cases in such a way to simplify our analysis and ensure that the cases we include in our study are the cases most likely to have died as a direct result from a cardiovascular disease. This is particularly important given the high prevalence of cardiovascular diseases.

The CDC's Wide-ranging ONline Data for Epidemiologic Research (WONDER) is a resource for a wide array of datasets surrounding the population health of the United States. There, you can find underlying cause of death data, which can be grouped by other characteristics such as sex, year, geographic region, and more. We will be examining diseases of the heart. These can be obtained using the following UCD-10 codes [here](https://wonder.cdc.gov/ucd-icd10.html):

+ I00-I09
+ I11
+ I13
+ I20-I51

# Data import

```{r, echo=FALSE}
# Temporary version of the OCS theme
theme_OCS <- theme(axis.text = element_text(size = 10, color="black",family = "Menlo", hjust = 0, face = "bold"),
        axis.line = element_line(size = 1, color = "black"),
        axis.title = element_text(size = 10, color="black",family = "Menlo", hjust = 0, face = "bold"),
        axis.ticks = element_line(size=1,color = "black"),
        panel.grid.major = element_line(color = "gray"),
        panel.grid.minor = element_line(color = "gray"),
        panel.border = element_blank(),
        panel.background = element_rect(fill=NA),
        plot.background = element_rect(fill = "#FFFFFF", color = "#FFFFFF"),
        plot.title = element_text(size = 14, color="black",family = "Menlo", hjust = 0, face = "bold"),
        plot.subtitle = element_text(color="black",family = "Menlo",size = 12, hjust = 0),
        legend.text = element_text(size=8,family = "Menlo", color="black"),
        legend.title = element_text(size=8, family = "Menlo", color="black", face = "bold", hjust = 0),
        legend.background = element_rect(fill="#FFFFFF", color = "#FFFFFF"),
        plot.caption = element_text(size=8, family = "Menlo", color="black", hjust = 0))
```

Let's import the data for this analysis. 

## United States county boundaries

We'll begin by importing the United States county boundaries data. Note that we are importing two different files. We must determine which of the two files to use. These two files have different resolutions.

The spatial data format we will be using for our analysis is known as the shapefile format. Shapefiles are a file format originated by Esri, the developers of the popular ArcGIS. Shapefiles can store a wide array of spatial data and are the most widely used format for spatial data. 

Using `read_sf` by the `sf` package, we can read in the shapefiles for our analysis. 

```{r}
map_sf <- read_sf("cb_2017_us_county_20m/cb_2017_us_county_20m.shp")
map_sf <- st_transform(map_sf, 2163)
# Need to show source for this document

map_sf_high_res <- read_sf("cb_2017_us_county_500k/cb_2017_us_county_500k.shp")

FIPS_codes <- read_excel("all-geocodes-v2017.xlsx")
```

Let's compare the boundaries of each file. 

If we plot each, we can see that there is substantially more detail in the larger of the two shapefiles. This could be useful when generating maps that are smaller in scale or when the resolution could impact an analysis.

```{r, echo=FALSE, warning=FALSE}
NY_shape <- map_sf %>% filter(STATEFP=="36" | STATEFP=="34" | STATEFP =="42")
NY_shape_high_res <- map_sf_high_res %>% filter(STATEFP=="36" | STATEFP=="34" | STATEFP =="42")

#There are 58 counties in California. Since three are shown, a warning regarding the other 55 is shown. 

NY_subset <- ggplot() +
  geom_sf(data=NY_shape, color="blue", fill="transparent", size = 0.5) +
  geom_sf(data=NY_shape_high_res, color="red", fill="transparent", size = 0.5) +
  geom_sf_text(data=NY_shape_high_res, aes(label = NAME), size = 2.5) +
  xlim(2125000, 2175000) + 
  ylim(-150000, -100000) +
  xlab("Longitude") +
  ylab("Latitude")+
  theme_void()
```

With that said, if we zoom out enough the two shapefiles are almost indiscernable from one another.

```{r, echo=FALSE, warning=FALSE}
NY_NJ_PA <- ggplot() + 
  geom_sf(data=NY_shape, color="blue", fill="transparent", size = 0.1) +
  geom_sf(data=NY_shape_high_res, color="red", fill="transparent", size = 0.1) +
  xlim(1592058, 2303517) + 
  ylim(-391238.6, 344335) +
  theme_void()

arrowA <- data.frame(x1 = 28, x2 = 35, y1 = 22, y2 = 47)
arrowB <- data.frame(x1 = 28, x2 = 35, y1 = 18, y2 = 3)

ggplot() +
    coord_equal(xlim = c(0, 70), ylim = c(0, 50), expand = FALSE) +
    annotation_custom(ggplotGrob(NY_NJ_PA), xmin=0, xmax = 35, ymin = 0, 
        ymax = 50) +
    annotation_custom(ggplotGrob(NY_subset), xmin=35, xmax = 70, ymin = 0, 
        ymax = 50) +
  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2), data = arrowA, lineend = "round") +
  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2), data = arrowB, lineend = "round") +
  labs(title = "Impact of resolution of spatial data on shape produced", subtitle = "Smaller shapefile in blue, larger shapefile in red") +
  theme_void()
```

We see that for certain applications the use of the larger shapefile is not only warranted, but necessary. Conversely, for other applications the use of the larger shapefile is not necessary.

Since we are not conducting an analysis that could introduce bias or cause misclassification as a result of the shapefilefile we use, we are free to use the smaller of the two shapefiles. The smaller shapefile requires less computational resources and produces almost identical maps at the state level and beyond, making it preferable for this analysis.

Often, analyzing data thoroughly can be severely hindered by the computational resources available. This is especially true with spatial applications of R. Unless an analysis requires that those computational resources be used, it is often more efficient and effective to proceed with an analysis in a way that is less computationally-intensive.

Again, because our analysis is not dependent on the absolute precision of the maps we use, but rather the relative location of boundaries, we are able to use the smaller, lower-resolution file. If, for example, our analysis involved the aggregation of points inside a county, we would not have this luxury; the lower resolution would likely lead to points near the true boundary being misclassified as being within the borders of a nearby county. 

## ACS data

Let's import the following datasets for 2017: 

TOTAL POPULATION
Universe: Total population  more information
2013-2017 American Community Survey 5-Year Estimates

[Click here to be directed to the address for download](https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_B01003&prodType=table)

SELECTED CHARACTERISTICS OF HEALTH INSURANCE COVERAGE IN THE UNITED STATES
2013-2017 American Community Survey 5-Year Estimates

[Click here to be directed to the address for download](https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_S2701&prodType=table)

MEDIAN INCOME IN THE PAST 12 MONTHS (IN 2017 INFLATION-ADJUSTED DOLLARS)
2013-2017 American Community Survey 5-Year Estimates

[Click here to be directed to the address for download](https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_S1903&prodType=table)

MEDIAN AGE BY SEX
Universe: Total population
2013-2017 American Community Survey 5-Year Estimates

[Click here to be directed to the address for download](https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_B01002&prodType=table)

RACE
Universe: Total population
2013-2017 American Community Survey 5-Year Estimates

[Click here to be directed to the address for download](https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_B02001&prodType=table)

HISPANIC OR LATINO ORIGIN BY RACE
Universe: Total population
2013-2017 American Community Survey 5-Year Estimates

[Click here to be directed to the address for download](https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_B03002&prodType=table)

These datasets will serve as predictors in the model we create. 

```{r}
total_pop <- read_csv("ACS_17_5YR_B01003/ACS_17_5YR_B01003_with_ann.csv")

med_age <- read_csv("ACS_17_5YR_B01002/ACS_17_5YR_B01002_with_ann.csv")

race <- read_csv("ACS_17_5YR_B02001/ACS_17_5YR_B02001_with_ann.csv")

ethnicity <- read_csv("ACS_17_5YR_B03002/ACS_17_5YR_B03002_with_ann.csv")

med_inc <- read_csv("ACS_17_5YR_S1903/ACS_17_5YR_S1903_with_ann.csv")

perc_insured <- read_csv("ACS_17_5YR_S2701/ACS_17_5YR_S2701_with_ann.csv")
```

## Underlying cause of death data

Let's import the underlying cause of death data. 

This dataset will provide data for the counts in our Poisson regression model .

```{r}
ucd_heart <- read_tsv("ucd_cardiovascular_diseases_county_2017.txt",
                      na = c("Suppressed","Unreliable", "Missing"))
glimpse(ucd_heart)
```

Now that we've imported the data, we can wrangle the data for our analysis needs. 

# Data wrangling

We cannot wrangle a dataset for a specific purpose without first developing a clear idea of its components. This is particularly true when there are many columns. 

The `head()` and `glimpse()` functions from the `utils` and `tidyverse` packages are particularly useful for that purpose. Each function has an advantage and disadvantage over the other. The `head()` function returns the first several rows of data; this output, while very easy to read for datasets with few dimensions, is too messy to interpret in the presence of more than a handful of columns of data. The `glimpse()` function returns the first several rows for each and every column of a dataframe; this output, while thorough, can generate a lot of output that is difficult to sift through.

Rather than use one exclusively, we will use whichever is more beneficial for the dataset we are working with.

Using the `dim()` function from `base`, we can make that determination. 

If a function has few dimensions, we will use the `head()` function. If a function has more than 5 columns, we will use the `glimpse()` function. If a function too many dimensions to sift through reasonably, we will use regular expressions to identify our variable of interest. 
## ACS data

We'll begin by wrangling the American Community Survey data. 

### Total population

Let's prepare the dataset with the total population for each county. 

```{r}
dim(total_pop)
```

This dataset has `r dim(total_pop)[2]` columns and `r dim(total_pop)[1]` rows. Given the small number of columns, it's more expedient to use the `head()` function to first visualize the dataframe. 

```{r}
head(total_pop)
```

A few things can be said about the dataframe from the above output. 

The column names of this dataset correspond to a codebook and are difficult to interpret on their own. The first row of the dataframe uses more interpretable column names. In addition to this, not all the columns have information we need for our analysis.

We need to obtain practical column names and keep the information we need, mainly the county and corresponding estimates. 

Below we do just that. We used the first row to replace the column names. Then, we rid ourselves of unnecessary columns. Alternatively, we could have used the `select()` function by `dplyr`, a part of the `tidyverse`, to rid ourselves of the unnecessary columns.  

```{r}
colnames(total_pop) <- total_pop[1,]
total_pop <- total_pop[-1,]
total_pop$Id <- NULL
total_pop$`Margin of Error; Total` <- NULL
total_pop$Geography <- NULL
```

If we look at our dataframe now, each population estimate is coupled to a five digit code. Each US county has a unique identifier, called a FIPS code, that allows tabular data to be joined to spatial data. 

```{r}
head(total_pop)
```

Eventually, we will be joining this dataframe to our spatial data. Rather than join our tabular data based upon the name of each county, we will use this five digit code to join our tabular data to our spatial data. This avoids having to use regular expressions to make the names of counties parsimonious before joining, making such joins both more efficient and **accurate**. 

*MICHAEL:LEAH* In the blurb above, I'm not sure if **accurate** is appropriate. Is there a word commonly used in data science that describes correctly attributing information to an observation/row. In much of the spatial work I've done, such joins are incredibly important. I think it'd be extremely beneficial here if we could highlight the importance of reference codes using data science language.

For wrangling and regression analysis purposes, we need to rename our columns. Our other dataframes, also from the American Community Survey, will also use unique FIPS codes to refer to counties. 

*MICHAEL*: there is a need to call this a FIPS code rather than GEOID. 

```{r}
total_pop <- total_pop %>%
  rename("GEOID"=Id2, "Population"=`Estimate; Total`)
```

### Median age

Let's wrangle our dataset representing the median age of each county. 

```{r}
dim(med_age)
```

This dataset, because of the number of columns it has (`r dim(med_age)[2]`), would be better examined with `glimpse()`. 

```{r}
glimpse(med_age)
```

Like the prior dataset, this dataset uses column names that correspond to a codebook. If we look closely, it appears that the first observation of each column, that is the first row, contains more practical names. Our wrangling approach is therefore pretty similar.

```{r}
colnames(med_age)
colnames(med_age) <- med_age[1,]
med_age <- med_age[-1,]
med_age$Id <- NULL
med_age <- med_age %>%
  rename("GEOID"=Id2, "median_age"=`Estimate; Median age -- - Total:`) %>%
  dplyr::select(GEOID, median_age)
class(med_age$median_age)
med_age$median_age <- as.numeric(med_age$median_age)
```

### Race 

By now, its apparent that each of the remaining datasets will require similar steps to wrangle. We should begin to think about creating a function that will help streamline the wrangling process. Such a function would take some time to create to ensure that it produces the output we intend.

If we had to wrangle many similarly sourced datasets, this could make the wrangling process much more efficient and could save hours of work that could be spent directly analyzing the data. If done properly, such a function could actually prevent human error that commonly emerges when tasks that involve repitition.

The problem with such a function is that knowledge of the steps involved in the wrangling process is required in advance. Without such knowledge, mistakes in our function could be compounded across our datasets. Since we only have a handful of datasets to wrangle, our time would be better spent manually wrangling each dataset. 

The following dataset provides a great example of how a function is only as good as the foresight involved in its creation. 

The dataset has `r dim(race)[2]` columns.

```{r}
dim(race)
```

Using `glimpse()`, it becomes apparent that this dataset also uses impractical column names which could be made more practical with replacement by names included in the first row. 

```{r}
glimpse(race)
```

We replace the impractical column names with more practical column names.

```{r}
colnames(race)
colnames(race) <- race[1,]
race <- race[-1,]
race$Id <- NULL
```

We need to create several predictors with information provided by this dataset. To do so, we'll use several `dplyr` verbs.

Before getting into detail, note that the code below generates an error. 

```{r, eval=FALSE}
race <- race %>%
  rename("GEOID"=Id2,
         "Total"=`Estimate; Total:`,
         "W_only"=`Estimate; Total: - White alone`,
         "BA_only"=`Estimate; Total: - Black or African American alone`,
         "A_only"=`Estimate; Total: - Asian alone`) %>%
  mutate(Perc_W_only=W_only/Total,
         Perc_BA_only=BA_only/Total,
         Perc_A_only=A_only/Total) %>%
  dplyr::select(GEOID,
         Perc_W_only,
         Perc_BA_only,
         Perc_A_only)
```

If we check the class of each column, we find that each of the columns we treat numerically in the above code are of character class. 

```{r}
class(race$`Estimate; Total:`)
class(race$`Estimate; Total: - White alone`)
class(race$`Estimate; Total: - Black or African American alone`)
class(race$`Estimate; Total: - Asian alone`)
```

Using the `mutate()` function, we can coerce these columns to `numeric` class. This lets us complete the calculations we originally intended. 

Below we do just that. First, we rename the columns for practicality. Second, we coerce the columns to `numeric` class. Then, we calculate the predictors we'll be using for our regression analysis. Finally, we eliminate information that would not be relevant to our analysis with `select()`.

```{r}
race <- race %>%
  rename("GEOID"=Id2,
         "Total"=`Estimate; Total:`,
         "W_only"=`Estimate; Total: - White alone`,
         "BA_only"=`Estimate; Total: - Black or African American alone`,
         "A_only"=`Estimate; Total: - Asian alone`) %>%
  mutate("Total"=as.numeric(Total),
         "W_only"=as.numeric(W_only),
         "BA_only"=as.numeric(BA_only),
         "A_only"=as.numeric(A_only)) %>%
  mutate(Perc_W_only=W_only/Total,
         Perc_BA_only=BA_only/Total,
         Perc_A_only=A_only/Total) %>%
  dplyr::select(GEOID,
         Perc_W_only,
         Perc_BA_only,
         Perc_A_only)
```

While a function to wrangle our datasets could have been created to identify and deal with such issues, it could have only been done so with advanced knowledge of the problems that may emerge throughout the wrangling process. In the real world, such problems are often difficult or impossible to identify in advance. Such functions are therefore only useful in certain circumstances where the datasets involved are numerous and fairly similar in structure.

### Ethnicity

This next dataset follows a similar process. Data from the American Community Survey is very well maintained and has technical documentation. The similarity in the wrangling process is a result of the similarity in the way the datasets are organized. 

Using a similar wrangling process, we make the column names more practical and then calculate predictors that will be used in our regression analysis.

```{r}
dim(ethnicity)

glimpse(ethnicity)

colnames(ethnicity)
colnames(ethnicity) <- ethnicity[1,]
ethnicity <- ethnicity[-1,]
ethnicity$Id <- NULL

class(ethnicity$`Estimate; Total:`)
class(ethnicity$`Estimate; Hispanic or Latino:`)

ethnicity <- ethnicity %>%
  rename("GEOID"=Id2,
         "Total"=`Estimate; Total:`,
         "Hispanic_Latino"=`Estimate; Hispanic or Latino:`) %>%
  mutate("Total"=as.numeric(Total),
         "Hispanic_Latino"=as.numeric(Hispanic_Latino)) %>%
  mutate(Perc_Hispanic_Latino=Hispanic_Latino/Total) %>%
  dplyr::select(GEOID,
         Perc_Hispanic_Latino)
```

### Median income

The next couple of datasets, while from the American Community Survey, would be even more difficult to wrangle effectively with a function. 

The wrangling process for this dataset, which will eventually represent the median income of each county, is much more complex. 

The dataset has `r dim(med_inc)[2]` columns. This is far too many columns to sift through efficiently or effectively. 

```{r}
dim(med_inc)
```

We need to learn a bit more about this dataset before we can continue. 

Like our prior ACS datasets, the first 50 columns appear to pertain to a codebook.

```{r}
colnames(med_inc)[1:50]
```

The first row appears to have more practical column names.

```{r}
head(med_inc)
```

We substitute the impractical column names with more practical column names. 

```{r}
colnames(med_inc) <- med_inc[1, ]
med_inc <- med_inc[-1,]
med_inc$Id <- NULL
```

Now that we have practical column names, we can use regular expressions to search through the `dim(med_inc)[2]` columns. 

`str_detect()` returns a logical vector, with a value of `TRUE` being returned for specified patterns found within a string. `R` can sum these `TRUE` values. This sum gives a count of how many values satisfied the logical statement. 

We can use this to determine which patterns are most successful at narrowing down our search. 

By condititioning on two statements, we are able to reduce the `r dim(med_inc)[2]` columns to `r sum(str_detect(colnames(med_inc), "[Mm]edian [Ii]ncome") & str_detect(colnames(med_inc),"[Ee]stimate"))`. 

**Michael**: Explain pattern used. 

```{r}
sum(str_detect(colnames(med_inc), "[Mm]edian [Ii]ncome"))
sum(str_detect(colnames(med_inc), "[Ee]stimate"))
sum(str_detect(colnames(med_inc), "[Mm]edian [Ii]ncome") &
      str_detect(colnames(med_inc),"[Ee]stimate"))
```

We can then determine what these`r sum(str_detect(colnames(med_inc), "[Mm]edian [Ii]ncome") & str_detect(colnames(med_inc),"[Ee]stimate"))` column are. 

```{r}
colnames(med_inc)[which(str_detect(colnames(med_inc), "[Mm]edian [Ii]ncome") & 
                          str_detect(colnames(med_inc), "[Ee]stimate"))]
```

From here we can determine which columns we need and proceed as we did with the prior datasets.

```{r}
med_inc <- med_inc %>%
  rename("GEOID"=Id2,
         "median_income"=`Median income (dollars); Estimate; Households`) %>%
  dplyr::select(GEOID,
         median_income)

class(med_inc$median_income)
med_inc$median_income <- as.numeric(med_inc$median_income)
```

### Percent insured

*MICHAEL* Must explain regular expression decisions. 

```{r}
dim(perc_insured)

colnames(perc_insured)[1:50]

head(perc_insured)[,1:5]

colnames(perc_insured) <- perc_insured[1, ]

perc_insured <- perc_insured[-1,]
perc_insured$Id <- NULL

sum(str_detect(colnames(perc_insured), "[Pp]ercent [Ii]nsured"))
sum(str_detect(colnames(perc_insured), "[Ee]stimate"))
sum(str_detect(colnames(perc_insured), "[Pp]ercent [Ii]nsured") &
      str_detect(colnames(perc_insured),"[Ee]stimate"))

colnames(med_inc)[which(str_detect(colnames(med_inc), "Pp]ercent [Ii]nsured") & 
                          str_detect(colnames(med_inc), "[Ee]stimate"))]

perc_insured <- perc_insured %>%
  rename("GEOID"=Id2,
         "perc_insured"=`Percent Insured; Estimate; Civilian noninstitutionalized population`) %>%
  dplyr::select(GEOID,
         perc_insured)

class(perc_insured$perc_insured)
perc_insured$perc_insured <- as.numeric(perc_insured$perc_insured)
```

## Underlying cause of death data

*MICHAEL*: Must make decisions to use `NULL` clear. It seems as if these decisions are arbitrary. 

```{r}
dim(ucd_heart)

glimpse(ucd_heart)

ucd_heart$Notes <- NULL
ucd_heart$Population <- NULL
ucd_heart$`Crude Rate` <- NULL
ucd_heart$`2013 Urbanization` <- as.factor(ucd_heart$`2013 Urbanization`)
glimpse(ucd_heart)
```

*According to [this](https://www.cdc.gov/nchs/data/series/sr_02/sr02_166.pdf) document, the ordering should be something along these lines*

```{r}
ucd_heart <- ucd_heart %>%
  rename("GEOID"=`County Code`)
levels(ucd_heart$`2013 Urbanization`)
ucd_heart$`2013 Urbanization` <- factor(ucd_heart$`2013 Urbanization`,
       levels = c("NonCore (Nonmetro)",
                  "Micropolitan (Nonmetro)",
                  "Small Metro",
                  "Medium Metro",
                  "Large Fringe Metro",
                  "Large Central Metro"),
       ordered = TRUE)

levels(ucd_heart$`2013 Urbanization`)

ucd_heart <- ucd_heart %>% mutate("2013 Urbanization Simplified"=fct_collapse(`2013 Urbanization`,
               large_metro=c("Large Fringe Metro",
                  "Large Central Metro"),
               small_medium_metro=c("Small Metro",
                  "Medium Metro"),
               non_metro=c("NonCore (Nonmetro)",
                  "Micropolitan (Nonmetro)")))

#recode factors to simplify
```

## United States county boundaries

*MICHAEL*: Why is EPSG 2163 used? 

```{r}
st_crs(map_sf)

map_sf <- st_transform(map_sf, 2163)

dim(map_sf)

glimpse(map_sf)
```

```{r}
map_sf <- map_sf %>%
  left_join(ucd_heart, by="GEOID") %>%
  left_join(total_pop, by="GEOID")

class(map_sf$Deaths)
class(map_sf$Population)
class(map_sf$`2013 Urbanization`)

map_sf$Population <- as.numeric(map_sf$Population)

map_sf <- map_sf %>%
  mutate("Crude Rate"=round((Deaths/Population)*1e+06,3))
```

```{r}
glimpse(FIPS_codes)

head(FIPS_codes)

colnames(FIPS_codes)
FIPS_codes[4,1:6]
colnames(FIPS_codes) <- FIPS_codes[4,]
FIPS_codes <- FIPS_codes[-(1:4),]

glimpse(FIPS_codes)

sapply(FIPS_codes[,1:(length(colnames(FIPS_codes)))], class) # Apply this to other section of the code where checking of column class is necessary. 

FIPS_codes <- sapply(FIPS_codes[,1:length(colnames(FIPS_codes))], as.factor)

class(FIPS_codes)

FIPS_codes <- as.data.frame(FIPS_codes)

sapply(FIPS_codes[,1:(length(colnames(FIPS_codes))-1)], summary)

colnames(FIPS_codes)

FIPS_codes <- FIPS_codes %>%
  dplyr::select(`Summary Level`,
         `State Code (FIPS)`,
         `County Code (FIPS)`,
         `Area Name (including legal/statistical area description)`) %>%
  rename("summary_level"=`Summary Level`,
         "state_code"=`State Code (FIPS)`,
         "county_code"=`County Code (FIPS)`,
         "area_name"=`Area Name (including legal/statistical area description)`) %>%
  sapply(as.character) %>%
  as.data.frame() %>% #Discuss how using sapply converted our dataframe into a matrix, what we are doing essentially is turning it back into a dataframe for manipulation
  mutate("GEOID"=paste0(state_code,county_code))

FIPS_codes_dfs <- FIPS_codes %>%
  group_split(summary_level)

FIPS_codes_dfs %>%
  lapply(`[`, 1, )

state_codes <- FIPS_codes_dfs[[2]]
glimpse(state_codes)
state_codes <- state_codes %>%
  dplyr::select(state_code, area_name) %>%
  rename("state_name"=area_name)
glimpse(state_codes)
county_codes <- FIPS_codes_dfs[[3]]
glimpse(county_codes)
county_codes <- county_codes %>%
  dplyr::select(-summary_level) %>%
  rename("county_name"=area_name)

# Mention how there is a heirarchy to these codes 

covariates <- county_codes %>%
  left_join(state_codes, by="state_code") %>%
  left_join(med_age, by="GEOID") %>%
  left_join(race, by="GEOID") %>%
  left_join(ethnicity, by="GEOID") %>%
  left_join(med_inc, by="GEOID") %>%
  left_join(perc_insured, by="GEOID")
```

*Now that we have the reference codes, we can update this section*

We can use information from the US Census Bureau to identify which codes correspond to the states of interest (click. [here](https://www.census.gov/geographies/reference-files/2017/demo/popest/2017-fips.html)).

```{r}
map_sf <- map_sf %>%
  left_join(covariates, by="GEOID")
```

```{r}
# Using "" around numbers to prevent 0 from disappearing from "09" later in the analysis. If the zero disappears, R does not recognize CT in maps or incorrectly classifies it later in the analysis. 

state_fp <- c("39", #OH
              "54", #WV
              "51", #VA
              "24", #MD
              "10", #DE
              "34", #NJ
              "42", #PA 
              "36", #NY
              "09", #CT
              "44", #RI
              "25", #MA
              "50", #VT
              "33", #NH
              "23" #ME
              )
```

```{r}
subset <- map_sf %>%
  filter(STATEFP %in% state_fp)
```

```{r}
contiguous_us <- map_sf %>% filter(!(STATEFP %in% c("02", #Alaska
                                                    "15", #Hawaii
                                                    "72")))
```

## Lag Covariates

We need to construct the neighborhood matrix in a way that does includes all the relationships between counties of interest, otherwise we will lose valuable information and could bias our results.

Let's create a neighborhood matrix that includes counties just outside of our region of interest. 

Five states border our region of interest. Let's include them in our neighborhood matrix.

```{r}
# "" around numbers for the same reason as mentioned before

surrounding_states <- c("37", #NC
                        "47", #TN
                        "21", #KY
                        "18", #IN
                        "26" #MI
                        )

matrix_counties <- c(state_fp, surrounding_states)

matrix_region <- map_sf %>%
  filter(STATEFP %in% matrix_counties)

matrix_region <- matrix_region %>%
  mutate(Region = ifelse(STATEFP %in% state_fp, "Region of Interest", "Surrounding States"))

```

Let's look at what we did visually. 

```{r}
ggplot(matrix_region) + 
  geom_sf(aes(fill=Region))
```

```{r}
# Would be neat to give an example of queen's vs rook's case matrices
matrix_region_sp <- as(matrix_region, 'Spatial')
nbrs_matrix_region_sp <- poly2nb(matrix_region_sp, queen = TRUE)
neighbors_sf <- as(nb2lines(nbrs_matrix_region_sp, coords = coordinates(matrix_region_sp)), 'sf')
neighbors_sf <- st_set_crs(neighbors_sf, st_crs(matrix_region))
```

```{r}
ggplot(matrix_region) + 
  geom_sf(aes(fill = Region), color=NA) +
  geom_sf(data = neighbors_sf, size=0.25, color="black") +
  ggtitle("Neighborhood Matrix", subtitle = "Queen's Case Adjacency")
```

#### Smoothing


With this matrix, we can create smoothed maps. Smoothing is the process of borrowing information from surrounding area-level units to create values with less variation.

Our regression will require that we create lagged variables, variables with values that are summaries of proximal regions. What we consider proximal is a matter of our analysis question.

Recall our adjacency matrix. We can borrow information from area-level units—in our case, counties—that are queen adjacent. Doing this would be considered smoothing.

If we smoothed nearest queen adjacent area-level units, we would be creating a first order lagged variable. 

Below we create a first order lagged variable.

```{r}
# First order lag

FFct_nb_w <- nb2listw(nbrs_matrix_region_sp,style="W", zero.policy = TRUE)
Wfull<-listw2mat(FFct_nb_w)

matrix_region$smoothed_pop <- Wfull %*% matrix_region$Population

ggplot(matrix_region) +
  geom_sf(aes(fill=smoothed_pop)) +
  scale_fill_gradientn(colours = viridis(10))
```

If we smoothed using the next nearest adjacent area-level units, we would be creating a second order lagged variable. 

The process for creating a second order lagged variable, and so on, is the same.

```{r}
# Second order lag 

nb_sec_lag_list <- nblag(nbrs_matrix_region_sp, maxlag = 2)
nb_sec_lag <- nb_sec_lag_list[[2]]

FFct_nb_w <- nb2listw(nb_sec_lag,style="W", zero.policy = TRUE)
Wfull<-listw2mat(FFct_nb_w)

matrix_region$smoothed_pop_sec_order <- Wfull %*% matrix_region$Population

ggplot(matrix_region) +
  geom_sf(aes(fill=smoothed_pop_sec_order)) +
  scale_fill_gradientn(colours = viridis(10))
```

We can also combine combine information from the nearest and next nearest adjacent area-level units. 

The code below uses the list of first and second order adjacent counties to do exactly that. 

```{r}
# Cumulative 1st and 2nd order
nb_cum_lag <- nblag_cumul(nb_sec_lag_list)

FFct_nb_w <- nb2listw(nb_cum_lag,style="W", zero.policy = TRUE)
Wfull<-listw2mat(FFct_nb_w)

matrix_region$smoothed_pop_cum_order <- Wfull %*% matrix_region$Population

ggplot(matrix_region) +
  geom_sf(aes(fill=smoothed_pop_cum_order)) +
  scale_fill_gradientn(colours = viridis(10))
```

If we take an extreme case, we can better visualize the implications such smoothing can have on our data. 

Below, we take an average of the first through tenth order adjacent counties. 

The end product is a map with much less variation. 

```{r}
# Extreme case

nb_ext_lag_list <- nblag(nbrs_matrix_region_sp, maxlag = 10)
nb_ext_cum_lag <- nblag_cumul(nb_ext_lag_list)

FFct_nb_w <- nb2listw(nb_ext_cum_lag,style="W", zero.policy = TRUE)
Wfull<-listw2mat(FFct_nb_w)

matrix_region$smoothed_pop_ext_cum_order <- Wfull %*% matrix_region$Population

ggplot(matrix_region) +
  geom_sf(aes(fill=smoothed_pop_ext_cum_order)) +
  scale_fill_gradientn(colours = viridis(10))
```

If we plot the difference of the smoothed population from the actual population, we see that there is systematic overestimation and underestimation throughout this map. This is because of the effect smoothing has on our values. The more counties we use to generate an average summary value for a county, the more the summary value is alike to the average of the entire region. The reduced variation has both a cost and a benefit. 

```{r}
matrix_region <- matrix_region %>%
  mutate(Bias = Population - smoothed_pop_ext_cum_order) %>%
  mutate(Bias_cat = cut(Bias, breaks = 7, include.lowest = TRUE, ordered_result = TRUE)) %>%
  mutate("Bias Class" = ifelse(sign(Bias)==-1, "Overestimation", ifelse(sign(Bias)==1, "Underestimation", "Neither"))) 

summary(matrix_region$Bias)

ggplot(matrix_region) +
  geom_sf(aes(fill=Bias_cat)) +
  scale_fill_manual(values=c(viridis(7)))

ggplot(matrix_region) +
  geom_sf(aes(fill=`Bias Class`)) +
  scale_fill_manual(values = c(viridis(2)))
```

We can think of this in a different way. What if we were instead analyzing data for the entire contiguous US? If we smoothed enough so that the entire northeast region had a very similar population per county, the county values would be very similar to the average of the region. This idea still holds when we are analyzing a smaller region: if we smooth values enough, the values we produce tend to reflect the average of the region we our analyzing.

```{r}
# This would be a great example to show an error if you don't remove the one value without population. The matrix algebra would not work! 

contiguous_us <- contiguous_us %>% filter(is.na(Population)==FALSE)
contiguous_us_sp <- as(contiguous_us, 'Spatial')
nbrs_contiguous_us_sp <- poly2nb(contiguous_us_sp, queen = TRUE)

nb_ext_lag_list_contiguous <- nblag(nbrs_contiguous_us_sp, maxlag = 10)
nb_ext_cum_lag_contiguous <- nblag_cumul(nb_ext_lag_list_contiguous)

A <- nb2listw(nb_ext_cum_lag_contiguous,style="W", zero.policy = TRUE)
Wfulle <-listw2mat(A)

contiguous_us$smoothed_pop_ext_cum_order <- Wfulle %*% contiguous_us$Population

ggplot(contiguous_us) +
  geom_sf(aes(fill=smoothed_pop_ext_cum_order), color="transparent") + 
  scale_fill_gradientn(colors = viridis(10))
```

Broadly speaking, smoothing makes values look more like the global average, thereby reducing the global variance, exchanging exactitude in a map for the easier visualization. Smoothing creates fictitious values that are more than often overestimates or understimates of the true value. Despite this, we can use smoothing to advance our analysis with proper caution. Smoothing can help produce more interpretable visualizations as well as lagged variables that can be used in both inferential and predictive models. 

Using the code we used earlier, we can create a function that produces a lagged covariate. This is more efficient that manually creating each variable with the same code and prevents errors. Conversely, errors in our function can be systematically applied to each of our covariates. Since we already have the code and are assured that it functions as we intend, this is not a problem.

```{r}
FFct_nb_w <- nb2listw(nbrs_matrix_region_sp,style="W", zero.policy = TRUE)
Wfull<-listw2mat(FFct_nb_w)

lag_creator <- function(vector){
  lag_vector <- Wfull %*% vector
  lag_vector
}

lag_creator_plus <- function(vector, lags=2){
  lag_order_list <- nblag(nbrs_matrix_region_sp, maxlag = lags)
  nb_cum_lag <- nblag_cumul(lag_order_list)
  lag_vector <- Wfull %*% vector
  lag_vector
}

sum(complete.cases(covariates)==FALSE)

covariates[which(complete.cases(covariates)==FALSE),]

matrix_region <- matrix_region %>%
  mutate("median_age_lag"=lag_creator(median_age),
         "Perc_W_only_lag"=lag_creator(Perc_W_only),
         "Perc_BA_only_lag"=lag_creator(Perc_BA_only),
         "Perc_A_only_lag"=lag_creator(Perc_A_only),
         "Perc_Hispanic_Latino_lag"=lag_creator(Perc_Hispanic_Latino),
         "median_income_lag"=lag_creator(median_income),
         "perc_insured_lag"=lag_creator(perc_insured),
         "Population_lag"=lag_creator(Population)
         )

matrix_region <- matrix_region %>%
  mutate("median_age_lag2"=lag_creator_plus(median_age),
         "Perc_W_only_lag2"=lag_creator_plus(Perc_W_only),
         "Perc_BA_only_lag2"=lag_creator_plus(Perc_BA_only),
         "Perc_A_only_lag2"=lag_creator_plus(Perc_A_only),
         "Perc_Hispanic_Latino_lag2"=lag_creator_plus(Perc_Hispanic_Latino),
         "median_income_lag2"=lag_creator_plus(median_income),
         "perc_insured_lag2"=lag_creator_plus(perc_insured),
         "Population_lag2"=lag_creator_plus(Population)
         )

```

# Exploratory data analysis 

We can explore our data in different ways now that we've merged our data to one of the maps we've imported. 

The great thing about using sf objects is that they work very well with the `tidyverse` and `ggplot` packages. The code required to visualize and summarize our spatial data is extremely similar, if not identical to that used in non-spatial applications. 

Let's create a map of the crude mortality rate of cardiovascular diseases across the contiguous United States to help identify trends beyond our region of interest. 

It appears that rural regions of the United States may have greater crude mortality rates. 

```{r}
ggplot(contiguous_us) +
  geom_sf(aes(fill=log(`Crude Rate`)), color="black", lwd=0.1) +
  scale_fill_gradientn(colours = viridis(5)) +
  labs(title = "Crude Rate", subtitle = "Contiguous United States", caption = "Calculated with UCD data from CDC WONDER") +
  theme_OCS
```

If we look at the Northeastern United States... *update*

```{r}
ggplot(subset) +
  geom_sf(aes(fill=`Crude Rate`), color="black", lwd=0.1) +
  scale_fill_gradientn(colours = viridis(10)) +
  ggtitle("Crude Rate", subtitle = "Northeastern United States") + 
  theme_OCS
```

```{r}
st_crs(matrix_region)

LC_metro1 <- matrix_region %>% filter(`2013 Urbanization`=="Large Central Metro") %>% mutate(index=row_number())

st_agr(matrix_region) <- "constant"

st_agr(LC_metro1) <- "constant"

cen_dist_matrix <- st_distance(x=st_centroid(matrix_region), y=st_centroid(LC_metro1), "Great Circle")

matrix_region <- matrix_region %>% mutate(cen_dist_m=apply(data.frame(cen_dist_matrix), 1,min),
                         index=apply(cen_dist_matrix, 1, function(matrix_row) which(matrix_row==min(matrix_row))))

LC_metro1 <- LC_metro1 %>% as.data.frame() %>% dplyr::select(index, county_name) %>% rename("nearest_LC_metro"=county_name)

matrix_region <- matrix_region %>% left_join(LC_metro1, by="index")
```

*eval=FALSE*
matrix_region <- matrix_region %>% mutate(index=st_nearest_feature(st_centroid(matrix_region), st_centroid(LC_metro1)))
*chunk ends*

Now that we have calculated all of our spatial covariates, we can remove counties that are not part of our region of interest. 

```{r}
matrix_region <- matrix_region %>% filter(Region =="Region of Interest")
```

# Data analysis 

```{r}
model0 <- glm(Deaths ~ 1, 
            family="poisson",
            offset = log(Population),
            data=matrix_region,
            na.action = na.exclude)

matrix_region <- matrix_region %>%
  mutate("residuals0"=residuals.glm(model0, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals0), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white", high="red")

matrix_region <- matrix_region[is.na(matrix_region$residuals0)==FALSE,]
nbrs_matrix_region_sp <- poly2nb(matrix_region, queen = TRUE)
FFct_nb_w_test <- nb2listw(nbrs_matrix_region_sp,style="W", zero.policy = TRUE)

moran_cor_m0 <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals0,order=5,method="I",style="W", zero.policy = TRUE)
```

```{r}
model1 <- glm(Deaths ~ median_age + 
              Perc_W_only +
              Perc_BA_only + 
              Perc_A_only +
              Perc_Hispanic_Latino + 
              median_income +
              perc_insured + 
              `2013 Urbanization`, 
            family="poisson",
            offset = log(Population),
            data=matrix_region,
            na.action = na.exclude)

model2 <- glm(Deaths ~ median_age + 
              median_age_lag + 
              Perc_W_only + 
              Perc_W_only_lag +
              Perc_BA_only + 
              Perc_BA_only_lag + 
              Perc_A_only +
              Perc_A_only_lag + 
              Perc_Hispanic_Latino + 
              Perc_Hispanic_Latino_lag + 
              median_income +
              median_income_lag +
              perc_insured +
              perc_insured_lag + 
              `2013 Urbanization`, 
            family="poisson",
            offset = log(Population),
            data=matrix_region,
            na.action = na.exclude)
```

The presence of multicollinearity means that there is correlation between variables. However, high correlation does not mean that there is multicollinearity. 

Let's produce some correlation matrices and compare them with computed variance inflation factors for our variables. 

```{r}
melt_cor1 <- as.data.frame(matrix_region) %>%
  dplyr::select(median_age,
         Perc_W_only,
         Perc_BA_only,
         Perc_A_only,
         Perc_Hispanic_Latino,
         median_income,
         perc_insured) %>%
  cor() %>%
  round(3) %>%
  reshape2::melt()

colnames(melt_cor1)

melt_cor1 %>%
  ggplot(aes(x=Var1, y=Var2, fill=value, label=value)) +
  geom_tile() +
  scale_fill_gradient2(low="red", mid="white", high="green") + 
  geom_text() + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(angle = 45, vjust = 1, hjust = 1)) +
  ggtitle("Correlation matrix, focal effects") + 
  xlab("") +
  ylab("")
```

```{r}
melt_cor2 <- as.data.frame(matrix_region) %>%
  dplyr::select(median_age_lag,
         Perc_W_only_lag,
         Perc_BA_only_lag,
         Perc_A_only_lag,
         Perc_Hispanic_Latino_lag,
         median_income_lag,
         perc_insured_lag) %>%
  cor() %>%
  round(3) %>%
  reshape2::melt()

colnames(melt_cor2)

melt_cor2 %>%
  ggplot(aes(x=Var1, y=Var2, fill=value, label=value)) +
  geom_tile() +
  scale_fill_gradient2(low="red", mid="white", high="green") + 
  geom_text() + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(angle = 45, vjust = 1, hjust = 1)) +
  ggtitle("Correlation matrix, lagged effects") + 
  xlab("") +
  ylab("")
```

```{r}
melt_cor3 <- as.data.frame(matrix_region) %>%
  dplyr::select(median_age,
         Perc_W_only,
         Perc_BA_only,
         Perc_A_only,
         Perc_Hispanic_Latino,
         median_income,
         perc_insured,
         median_age_lag,
         Perc_W_only_lag,
         Perc_BA_only_lag,
         Perc_A_only_lag,
         Perc_Hispanic_Latino_lag,
         median_income_lag,
         perc_insured_lag) %>%
  cor() %>%
  round(3) %>%
  reshape2::melt()

colnames(melt_cor3)

melt_cor3 %>%
  ggplot(aes(x=Var1, y=Var2, fill=value, label=value)) +
  geom_tile() +
  scale_fill_gradient2(low="red", mid="white", high="green") + 
  geom_text(size=2) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(angle = 45, vjust = 1, hjust = 1)) +
  ggtitle("Correlation matrix, focal/lagged effects") + 
  xlab("") +
  ylab("")
```

```{r}
vif(model1)
vif(model2)
```

```{r}
sum(is.na(matrix_region$Deaths))
length(model1$residuals)
length(model2$residuals)
```

```{r}
#Are the residuals being plotted correctly? 

matrix_region <- matrix_region %>%
  mutate("residuals1"=residuals.glm(model1, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals1), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white", high="red")

matrix_region <- matrix_region %>%
  mutate("residuals2"=residuals.glm(model2, "response"))

matrix_region %>% 
  ggplot() +
  geom_sf(aes(fill=residuals2), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white", high="red")

moran_cor_m1 <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals1,order=5,method="I",style="W", zero.policy = TRUE)
moran_cor_m2 <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals2,order=5,method="I",style="W", zero.policy = TRUE)
plot(moran_cor_m1,ylim=c(-0.2, 0.4),main="Residuals from model 1")
plot(moran_cor_m2,ylim=c(-0.2, 0.4),main="Residuals from model 2")
```

matrix_region <- matrix_region[is.na(matrix_region$residuals1)==FALSE,]
nbrs_matrix_region_sp <- poly2nb(matrix_region, queen = TRUE)
FFct_nb_w_test <- nb2listw(nbrs_matrix_region_sp,style="W", zero.policy = TRUE)

Checking to see if average of neighbors is used. *This could be a great way to explain the weighted matrix. The code used in this case study appears somewhat black box-ish. It's important to not only describe the programming, but what it's actually doing.*

*This code does not work after the inclusion of code that removes surrounding states from the spatial data and was made no longer a chunk. An adaptation will be very helpful.*

matrix_region <- matrix_region %>% mutate(ID=row_number())
matrix_region %>% filter(GEOID==36047) %>% pull(median_income_lag)
matrix_region %>% filter(GEOID==36047) %>% pull(ID)
matrix_region %>% filter(ID %in% FFct_nb_w_test[[2]][[773]]) %>% dplyr::select(median_income, GEOID) %>% summarise(average=mean(median_income))


FFct_nb_w_test[[2]][[773]]
#Code checks out! 

```{r}
chi_square_model1 = sum(residuals(model1, type = "pearson")^2)

degrees_freedom_model1 = model1$df.residual

p_value_model1 = 1-pchisq(chi_square_model1, degrees_freedom_model1)

paste("Pearson goodness of fit = ", round(chi_square_model1,2), ", df = ", degrees_freedom_model1, ", p-value = ", round(p_value_model1,10))
```

```{r}
chi_square_model2 = sum(residuals(model2, type = "pearson")^2)

degrees_freedom_model2 = model1$df.residual

p_value_model2 = 1-pchisq(chi_square_model2, degrees_freedom_model2)

paste("Pearson goodness of fit = ", round(chi_square_model2,2), ", df = ", degrees_freedom_model2, ", p-value = ", round(p_value_model2,10))
```

Both models show overdispersion. 

```{r}
model1_nb <- glm.nb(Deaths ~ median_age +
                      Perc_W_only +
                      Perc_BA_only +
                      Perc_A_only + 
                      Perc_Hispanic_Latino + 
              median_income +
              perc_insured + 
              `2013 Urbanization` + 
              offset(log(Population)), 
            data=matrix_region,
            na.action = na.exclude)

model2_nb <- glm.nb(Deaths ~ median_age + 
              median_age_lag + 
              Perc_W_only + 
              Perc_W_only_lag +
              Perc_BA_only + 
              Perc_BA_only_lag + 
              Perc_A_only +
              Perc_A_only_lag + 
              Perc_Hispanic_Latino + 
              Perc_Hispanic_Latino_lag + 
              median_income +
              median_income_lag +
              perc_insured +
              perc_insured_lag +
              `2013 Urbanization` +
              offset(log(Population)),
            data=matrix_region,
            na.action = na.exclude)
```

```{r}
vif(model1_nb)
vif(model2_nb)
```

```{r}
matrix_region <- matrix_region %>%
  mutate("residuals1nb"=residuals.glm(model1_nb, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals1nb), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white" ,high="red")

matrix_region <- matrix_region %>%
  mutate("residuals2nb"=residuals.glm(model2_nb, "response"))

matrix_region %>% 
  ggplot() +
  geom_sf(aes(fill=residuals2nb), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white", high="red")

moran_cor_m1nb <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals1nb,order=5,method="I",style="W", zero.policy = TRUE)
moran_cor_m2nb <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals2nb,order=5,method="I",style="W", zero.policy = TRUE)

par(mfrow=c(2,2))

plot(moran_cor_m1,ylim=c(-0.2, 0.4),main="Residuals from model 1")
plot(moran_cor_m2,ylim=c(-0.2, 0.4),main="Residuals from model 2")
plot(moran_cor_m1nb,ylim=c(-0.2, 0.4),main="Residuals from model 1, nb")
plot(moran_cor_m2nb,ylim=c(-0.2, 0.4),main="Residuals from model 2, nb")
```

```{r}
model3 <- glm(Deaths ~ median_age + 
              Perc_W_only + 
              Perc_A_only +
              Perc_Hispanic_Latino + 
              median_income +
              perc_insured + 
              `2013 Urbanization`, 
            family="poisson",
            offset = log(Population),
            data=matrix_region,
            na.action = na.exclude)

model4 <- glm(Deaths ~ median_age + 
              median_age_lag + 
              Perc_W_only + 
              Perc_W_only_lag +
              Perc_A_only + 
              Perc_A_only_lag + 
              Perc_Hispanic_Latino + 
              Perc_Hispanic_Latino_lag + 
              median_income +
              median_income_lag +
              perc_insured +
              perc_insured_lag + 
                `2013 Urbanization`, 
            family="poisson",
            offset = log(Population),
            data=matrix_region,
            na.action = na.exclude)
```

```{r}
matrix_region <- matrix_region %>%
  mutate("residuals3"=residuals.glm(model3, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals3), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white" ,high="red")

matrix_region <- matrix_region %>%
  mutate("residuals4"=residuals.glm(model4, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals4), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white" ,high="red")

moran_cor_m3 <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals3,order=5,method="I",style="W", zero.policy = TRUE)
moran_cor_m4 <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals4,order=5,method="I",style="W", zero.policy = TRUE)
```

Does the next set of models exhibit overdispersion?

```{r}
chi_square_model3 = sum(residuals(model3, type = "pearson")^2)

degrees_freedom_model3 = model3$df.residual

p_value_model3 = 1-pchisq(chi_square_model3, degrees_freedom_model3)

paste("Pearson goodness of fit = ", round(chi_square_model3,2), ", df = ", degrees_freedom_model3, ", p-value = ", round(p_value_model3,10))
```

```{r}
chi_square_model4 = sum(residuals(model4, type = "pearson")^2)

degrees_freedom_model4 = model4$df.residual

p_value_model4 = 1-pchisq(chi_square_model4, degrees_freedom_model4)

paste("Pearson goodness of fit = ", round(chi_square_model4,2), ", df = ", degrees_freedom_model4, ", p-value = ", round(p_value_model4,10))
```

```{r}
model3_nb <- glm.nb(Deaths ~ median_age + 
              Perc_W_only + 
              Perc_A_only + 
              Perc_Hispanic_Latino + 
              median_income +
              perc_insured + 
              `2013 Urbanization` +
                offset(log(Population)),
            data=matrix_region,
            na.action = na.exclude)

model4_nb <- glm.nb(Deaths ~ median_age + 
              median_age_lag + 
              Perc_W_only + 
              Perc_W_only_lag + 
              Perc_A_only + 
              Perc_A_only_lag + 
              Perc_Hispanic_Latino + 
              Perc_Hispanic_Latino_lag + 
              median_income +
              median_income_lag +
              perc_insured +
              perc_insured_lag + 
              `2013 Urbanization` +
                offset(log(Population)),
            data=matrix_region,
            na.action = na.exclude)
```

```{r}
vif(model3_nb)
vif(model4_nb)
```

```{r}
matrix_region <- matrix_region %>%
  mutate("residuals3nb"=residuals.glm(model3_nb, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals3nb), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white" ,high="red")

matrix_region <- matrix_region %>%
  mutate("residuals4nb"=residuals.glm(model4_nb, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals4nb), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white" ,high="red")

moran_cor_m3nb <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals3nb,order=5,method="I",style="W", zero.policy = TRUE)
moran_cor_m4nb <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals4nb,order=5,method="I",style="W", zero.policy = TRUE)

par(mfrow=c(2,2))

plot(moran_cor_m3,ylim=c(-0.2, 0.4),main="Residuals from model 3")
plot(moran_cor_m4,ylim=c(-0.2, 0.4),main="Residuals from model 4")
plot(moran_cor_m3nb,ylim=c(-0.2, 0.4),main="Residuals from model 3, nb")
plot(moran_cor_m4nb,ylim=c(-0.2, 0.4),main="Residuals from model 4, nb")
```

```{r}
model5 <- glm(Deaths ~ median_age + 
              median_age_lag2 + 
              Perc_W_only + 
              Perc_W_only_lag2 +
              Perc_A_only + 
              Perc_A_only_lag2 + 
              Perc_Hispanic_Latino + 
              Perc_Hispanic_Latino_lag2 + 
              median_income +
              median_income_lag2 +
              perc_insured +
              perc_insured_lag2 + 
                `2013 Urbanization`, 
            family="poisson",
            offset = log(Population),
            data=matrix_region,
            na.action = na.exclude)

matrix_region <- matrix_region %>%
  mutate("residuals5"=residuals.glm(model5, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals5), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white" ,high="red")
```

```{r}
chi_square_model5 = sum(residuals(model5, type = "pearson")^2)

degrees_freedom_model5 = model5$df.residual

p_value_model5 = 1-pchisq(chi_square_model5, degrees_freedom_model5)

paste("Pearson goodness of fit = ", round(chi_square_model5,2), ", df = ", degrees_freedom_model5, ", p-value = ", round(p_value_model5,10))
```

```{r}
model5_nb <- glm.nb(Deaths ~ median_age + 
              median_age_lag2 + 
              Perc_W_only + 
              Perc_W_only_lag2 + 
              Perc_A_only + 
              Perc_A_only_lag2 + 
              Perc_Hispanic_Latino + 
              Perc_Hispanic_Latino_lag2 + 
              median_income +
              median_income_lag2 +
              perc_insured +
              perc_insured_lag2 + 
              `2013 Urbanization` +
                offset(log(Population)),
            data=matrix_region,
            na.action = na.exclude)

matrix_region <- matrix_region %>%
  mutate("residuals5nb"=residuals.glm(model5_nb, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals5nb), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white" ,high="red")
```

```{r}
moran_cor_m5 <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals5,order=5,method="I",style="W", zero.policy = TRUE)
moran_cor_m5nb <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals5nb,order=5,method="I",style="W", zero.policy = TRUE)

par(mfrow=c(1,2))

plot(moran_cor_m5,ylim=c(-0.2, 0.4),main="Residuals from model 5")
plot(moran_cor_m5nb,ylim=c(-0.2, 0.4),main="Residuals from model 5, nb")
```

```{r}
model6 <- glm(Deaths ~ median_age + 
              median_age_lag2 + 
              Perc_W_only + 
              Perc_W_only_lag2 +
              Perc_A_only + 
              Perc_A_only_lag2 + 
              Perc_Hispanic_Latino + 
              Perc_Hispanic_Latino_lag2 + 
              median_income +
              median_income_lag2 +
              perc_insured +
              perc_insured_lag2 + 
                `2013 Urbanization`,
              Population_lag,
            family="poisson",
            offset = log(Population),
            data=matrix_region,
            na.action = na.exclude)

matrix_region <- matrix_region %>%
  mutate("residuals6"=residuals.glm(model6, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals6), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white" ,high="red")

moran_cor_m6 <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals6,order=5,method="I",style="W", zero.policy = TRUE)
```

```{r}
chi_square_model6 = sum(residuals(model6, type = "pearson")^2)

degrees_freedom_model6 = model6$df.residual

p_value_model6 = 1-pchisq(chi_square_model6, degrees_freedom_model6)

paste("Pearson goodness of fit = ", round(chi_square_model6,2), ", df = ", degrees_freedom_model6, ", p-value = ", round(p_value_model6,10))
```

```{r}
#not the same as model6
model6_nb <- glm.nb(Deaths ~ median_age + 
                      median_age_lag +
              Perc_W_only + 
              Perc_W_only_lag + 
              median_income +
              median_income_lag + 
              perc_insured +
              perc_insured_lag +
                cen_dist_m +
                nearest_LC_metro +
                `2013 Urbanization Simplified` + 
                offset(log(Population)),
            data=matrix_region,
            na.action = na.exclude)

matrix_region <- matrix_region %>%
  mutate("residuals6nb"=residuals.glm(model6_nb, "response"))

matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=residuals6nb), color="transparent") + 
  scale_fill_gradient2(low="green", mid="white" ,high="red")

moran_cor_m6nb <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals6nb,order=5,method="I",style="W", zero.policy = TRUE)

plot(moran_cor_m6,ylim=c(-0.2, 0.4),main="Residuals from model 6")
plot(moran_cor_m6nb,ylim=c(-0.2, 0.4),main="Residuals from model 6, nb")
```

```{r}
matrix_region %>%
  ggplot() +
  geom_sf(aes(fill=cen_dist_m), color="transparent") + 
  scale_fill_gradient2(low="white" ,high="black")
```

```{r}
vif(model6_nb)
```

```{r}
ggplot(matrix_region, aes(x=Population, y=residuals6nb, label=County, color=`2013 Urbanization`)) +
  geom_point() + 
  geom_text()
```

```{r}
morans_i <- c(moran_cor_m1nb["res"][[1]][,1],
      moran_cor_m2nb["res"][[1]][,1],
      moran_cor_m3nb["res"][[1]][,1],
      moran_cor_m4nb["res"][[1]][,1],
      moran_cor_m5nb["res"][[1]][,1],
      moran_cor_m6nb["res"][[1]][,1]
      )

model <- c(rep("M1NB",5),
           rep("M2NB",5),
           rep("M3NB",5),
           rep("M4NB",5),
           rep("M5NB",5),
           rep("M6NB",5)
           )

lag <- rep(seq(1,5,by=1),6)

cor_df <- as.data.frame(cbind(morans_i, model, lag))

cor_df$morans_i <- as.numeric(as.character(cor_df$morans_i))

cor_df$model <- factor(cor_df$model, levels=c("M1NB",
                                                 "M2NB",
                                                 "M3NB",
                                                 "M4NB",
                                                 "M5NB",
                                              "M6NB"))

cor_df$lag <- as.numeric(as.character(cor_df$lag))

cor_df %>%
  ggplot(aes(x=lag, y=morans_i, color=model)) +
  geom_point(size=2) + 
  geom_line() + 
  geom_hline(yintercept = 0, color="black") + 
  ggtitle("Moran's I Correlogram by Model", subtitle = "Our models do not account for the spatial autocorrelation of our autocome.") + 
  facet_grid(.~model) + 
  xlim(1,5) + 
  ylim(-0.2,0.4) + 
  theme_OCS
```

We need to be careful about over-interpreting correlograms, particularly when the degree of autocorrelation is very small. 

We can test for autocorrelation globally by estimating the Moran's I statistic

```{r}
moran.mc(matrix_region$residuals6nb, listw =FFct_nb_w_test,nsim = 9999, zero.policy = TRUE)
```

Our Moran's I statistic is close to 0 and is not significant. Our model appears to have accounted for residual spatial autocorrelation.

Residual spatial autocorrelation can have a large impact on the usefulness of our model. Had we not accounted for it, our standard errors would have been underestimated, amongst other things. Ultimately, such behavior in the residuals is indicative of unaccounted for information in a model.

matrix_region <- matrix_region %>%
  mutate("residuals7nb"=residuals.glm(model7_nb, "response"))

moran_cor_m7nb <-sp.correlogram(neighbours=FFct_nb_w_test$neighbours,var=matrix_region$residuals7nb,order=5,method="I",style="W", zero.policy = TRUE)

plot(moran_cor_m6nb,ylim=c(-0.2, 0.4),main="Residuals from model 6, nb")
plot(moran_cor_m7nb,ylim=c(-0.2, 0.4),main="Residuals from model 7, nb")



moran.mc(matrix_region$residuals6nb, listw =FFct_nb_w_test,nsim = 9999, zero.policy = TRUE)


# Summary of results

# Code garage

```{r, eval=FALSE}
tm_shape(subset) +
	tm_polygons(c("Deaths", "Population", "Crude Rate", "2013 Urbanization"),
				style=c("pretty","pretty","pretty", "pretty"), breaks=list(NULL, NULL, NULL, NULL),
				palette=list("Oranges", "Purples", "Reds", "Blues"),
				border.col = "white",
				title=c("Deaths", "Population", "Crude Rate", "2013 Urbanization")) + tm_facets(as.layers = TRUE)
```
				
# Downloaded data

Total population 5 year 2017
https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_B01003&prodType=table

https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_S2701&prodType=table
SELECTED CHARACTERISTICS OF HEALTH INSURANCE COVERAGE IN THE UNITED STATES  more information
2013-2017 American Community Survey 5-Year Estimates

https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_S1903&prodType=table
MEDIAN INCOME IN THE PAST 12 MONTHS (IN 2017 INFLATION-ADJUSTED DOLLARS)  more information
2013-2017 American Community Survey 5-Year Estimates

https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_1YR_B01002&prodType=table
MEDIAN AGE BY SEX
Universe: Total population  more information
2017 American Community Survey 1-Year Estimates

https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_B02001&prodType=table
RACE
Universe: Total population  more information
2013-2017 American Community Survey 5-Year Estimates

https://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_17_5YR_B03002&prodType=table
HISPANIC OR LATINO ORIGIN BY RACE
Universe: Total population  more information
2013-2017 American Community Survey 5-Year Estimates

UCD Death Data—CDC Wonder
https://wonder.cdc.gov/ucd-icd10.html


pivot wide pivot longr in tidyr for reshape melt, change font theme

reinforce weight matrix multiple to average--Leah Meeting 

check for truncation effect of urban centers

explanation of edge vs centroid.


# Further Reading

[Residual spatial autocorrelation in macroecological and biogeographical modeling: a review](https://link.springer.com/article/10.1186/s41610-019-0118-3)
